Higher-Order unification for free!

### 2024-04-12 paper/main.tex
up to 3.5

Unification variables are identified by a
natural number that represents a memory addresses.
The memory and its associated operations are described below:

@@

Since each occurrence of a \Ho unification variables has a scope,
its assignment needs to be abstracted over it to enable the
instantiation of the same assignment to different scopes.
This is expressed by the \elpiIn{inctx} container, and in particular
its \elpiIn{abs} binding constructor.
On the contrary a solution to a \Fo variable is a plain term.

\input{code/fo_subst}
\input{code/ho_subst}

\noindent
We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
the one of \Ho.

@@

Each \elpiIn{hvariable} is stored in the mapping together with
its arity so that the code of \ref{clause:malloc} below can preserve:

@@

Applying the substitution corresponds to dereferencing a term with respect to
the memory.
%  To ease the comparison we split \Fo{} dereferencing into a
% \elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
% to memory cells that are set with their values, and has a corresponding
% operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
% has no corresponding operation in \Ho, and only ensures that
% terms of the form <<\elpiIn{fapp[fapp L1|L2]}>> are replaced by
% <<\elpiIn{fapp L3}>> where \elpiIn{L3} is the concatenation of \elpiIn{L1}
% and \elpiIn{L2}. The reasons for this asymmetry is
% that an \elpiIn{fapp} node with a flexible head is always mapped
% to a \elpiIn{uva} (as per \cref{sec:simulation,sec:beta}),
% preventing nested applications to materialize.
% \input{code/fderef}
It is worth looking at the code fpr \Ho{} to remark how assignments are moved
to the current scope, i.e. renaming the \elpiIn{abs}-bound variables
with the names in the scope of the unification variable occurrence.

@@

\noindent
Note that move strongly relies on invariant \ref{inv:uvaarity}: the length
of the arguments of all occurrences of a unification are the same. Hence
they have the same type in the meta-level and the number of \elpiIn{abs}
nodes in the assignment matches that length. In turn this grants that
\elpiIn{move} never fails.\todo{convoluted}

@@

\noindent
The right hand side of a link, the problematic term, can occur under binders.
To accommodate this situation the compiler wraps \elpiIn{baselink} using
the \elpiIn{inctx} container (see rule~\ref{data:inctx}).

@@

\noindent
New links are suspended by construction.
If the left hand side variable is assigned during a step, then 
the link is considered for progress and possibly eliminated.
This is discussed in \cref{sec:eta} and \cref{sec:beta}.

@@

\subsection{Compilation}
\label{sec:compilation}
\noteE[inline]{manca beta normal in entrata}

The main task of the compiler is to recognize \Fo{} variables standing
for functions and map them to higher order variables in \Ho.
In order to bring back the substitution from \Ho{} to \Fo{} the compiler
builds a ``memory map'' connecting the the kind of variables using routine
\ref{clause:malloc}.

The signature of the \elpiIn{comp} predicate below allows for the generation of
links (suspended unification problems) that play no role in this section
but play a major role in \cref{sec:eta} and \cref{sec:beta}.
With respect to \cref{sec:problem-statement} the signature also allows
for updates to the substitution.

\input{code/comp_base}

@@

\noindent
Note that we could remove the first rule, whose solve purpose is to make
links more readable by pruning unused context entries.

@@

A step in \Ho consists in unifying two terms and reconsidering all
links for progress. If any of the two tasks fail we say that the entire step
fails, and it is at this granularity that we can relate steps in the
two languages.

@@

\subsubsection{Progress}
In the base compilation scheme \elpiIn{progress1} is the identity
on both the links and the substitution, so the fixpoint trivially terminates.
Sections \ref{sec:eta} and \ref{sec:beta} add rules to \elpiIn{progress1}
and justify why the don't hinder termination. For brevity we omit the code
that applies the substitution \elpiIn{S1} to all terms in \linkStore.

\subsubsection{Occur check}
Since compilation moves problematic terms out of the sigh of \Ue{},
that procedure can only perform a partial occur check. For example the
unification problem $X \Ue f~Y$ cannot generate a cyclic substitution alone,
but should be disallowed if a $\linkStore$ contains a link like
$\linketaM{}{Y}{\lambda z.X_z}$: We don't know yet if $Y$ will feature
a lambda in head position, but we surely know it contains $X$, hence
$f~Y$ and that fails the occur check.
The procedure \elpiIn{occur-check-links} is in charge of ensuring that
each link does not represent a (suspended) unification problem doomed
to fail because of occur check. This check is needed in order to
guarantee \fullref{prop:fidelity}.

>>>>

Unification variables are identified by a (unique) memory address.
The memory and its associated operations are described below:

@@

Since each \Ho unification variables occurs together with a scope,
its assignment needs to be abstracted over it to enable the
instantiation of the same assignment to different scopes.
This is expressed by the \elpiIn{inctx} container, and in particular
its \elpiIn{abs} binding constructor.

\input{code/ho_subst}

\noindent
A solution to a \Fo{} variable is a plain term, that is \elpiIn{fsubst}
is an abbreviation for \elpiIn{mem fm}.
%We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
%the one of \Ho.

@@

\noindent
Each \elpiIn{hvariable} is stored in the mapping together with
its arity (a number) so that the code of \ref{clause:malloc} below can preserve:

@@

%Applying the substitution corresponds to dereferencing a term with respect to
%the memory.
%  To ease the comparison we split \Fo{} dereferencing into a
% \elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
% to memory cells that are set with their values, and has a corresponding
% operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
% has no corresponding operation in \Ho, and only ensures that
% terms of the form <<\elpiIn{fapp[fapp L1|L2]}>> are replaced by
% <<\elpiIn{fapp L3}>> where \elpiIn{L3} is the concatenation of \elpiIn{L1}
% and \elpiIn{L2}. The reasons for this asymmetry is
% that an \elpiIn{fapp} node with a flexible head is always mapped
% to a \elpiIn{uva} (as per \cref{sec:simulation,sec:beta}),
% preventing nested applications to materialize.
% \input{code/fderef}
It is worth looking at the code of \elpiIn{deref} that
applies the substitution to a \Ho{} term. Remark how assignments are moved
to the current scope, i.e. renaming the \elpiIn{abs}-bound variables
with the names in the scope of the unification variable occurrence.

@@

\noindent
Note that move strongly relies on invariant \ref{inv:uvaarity}: the length
of the arguments of all occurrences of a unification are the same. Hence
they have the same simple type for the meta-level and hence the number of
\elpiIn{abs} nodes in the assignment matches that length.
In turn this grants that \elpiIn{move} never fails.

@@

\noindent
The right hand side of a link, the problematic term, can occur under binders.
To accommodate this situation the compiler wraps \elpiIn{baselink} using
the \elpiIn{inctx} container (see~\ref{data:inctx} also used for \elpiIn{subst}).

@@

\noindent
New links are suspended by construction.
If the left hand side is assigned during a step, then 
the link is considered for progress and possibly eliminated.
This is discussed in \cref{sec:eta} and \cref{sec:beta}.

@@

\subsection{Compilation}
\label{sec:compilation}

The simple compiler described in this section serves as a base for the
extensions in \cref{sec:llam,sec:eta,sec:beta}.
Its main task is to beta normalize the term and map one syntax tree to
the other.
In order to bring back the substitution from \Ho{} to \Fo{} the compiler
builds a ``memory map'' connecting the the kind of variables using routine
\ref{clause:malloc}.

The signature of the \elpiIn{comp} predicate below allows for the generation of
links (suspended unification problems) that play no role in this section
but play a major role in \cref{sec:llam,sec:eta,sec:beta}.
With respect to \cref{sec:problem-statement} the signature also allows
for updates to the substitution.

\input{code/comp_base}
\todo{use the real code}
\begin{elpicode}
type compile fm -> tm -> mmap -> mmap -> links -> links ->
  subst -> subst -> o.
compile Fo0 Ho M0 M1 L0 M1 S0 S1 :-
  beta-normal Fo0 Fo, comp Fo Ho M0 M1 L0 M1 S0 S1.
\end{elpicode}

@@

\noindent
Note that we could remove the first rule, whose sole purpose is to make
links more readable by pruning unused context entries.

@@

A step in \Ho consists in unifying two terms and reconsidering all
links for progress. If any of the two tasks fails we say that the entire step
fails, and it is at this granularity that we can relate steps in the
two languages.

@@

\subsubsection{Progress}
In the base compilation scheme \elpiIn{progress1} is the identity
on both the links and the substitution, so the fixpoint trivially terminates.
Sections \ref{sec:eta} and \ref{sec:beta} add rules to \elpiIn{progress1}
and justify why the don't hinder termination.
% For brevity we omit the code
% that applies the substitution \elpiIn{S1} to all terms in \linkStore.

\subsubsection{Occur check}
Since compilation moves problematic terms out of the sight of \Ue{},
that procedure can only perform a partial occur check. For example the
unification problem $X \Ue f~Y$ cannot generate a cyclic substitution alone,
but should be disallowed if a $\linkStore$ contains a link like
$\linketaM{}{Y}{\lambda z.X_z}$: we don't know yet if $Y$ will feature
a lambda in head position, but we surely know it contains $X$, hence
$f~Y$ and that fails the occur check.
The procedure \elpiIn{occur-check-links} is in charge of
performing this check that is needed in order to
guarantee \fullref{prop:fidelity}.
